---
title: "Study on the prediction of the ... "
author: "Thawatchai Phakwithoonchai"
date: "2/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(caret)
library(corrplot)
library(gbm)
library(ggplot2)
library(rattle)
library(randomForest)
```

# Executive Summary  
The objective of study is to predict the manner in which 6 participants did the exercise. After getting, cleaning, and performing the data preparation. The models were built based on 3 algorithm that were (1) Classification Trees: CART, (2) Random Forests: RF, and (3) Generalized Boosted Regression Modeling: GBM. 

# Getting, Cleaning, and Data Preparation  
Training dataset was then seperated into 2 groups, which were (1) training set (70% of the training data), and (2) testing set (the 30% of remaining data), while the validation data was still unchanged.

``` {r loading and exploring the data}
trainURL = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testURL = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(trainURL, "./training.csv")
download.file(testURL, "./testing.csv")
raw.training <- read.csv("./training.csv")
validation <- read.csv("./testing.csv")
t(data.frame(Training = dim(raw.training), Validation = dim(validation), row.names = c("Observations", "Variables")))
```

``` {r separating the training data}
set.seed(12345)
inTrain <- createDataPartition(y = raw.training$classe, p = 0.7, list = FALSE)
training <- raw.training[inTrain, ]
testing <- raw.training[-inTrain, ]
t(data.frame(Training = dim(training), Testing = dim(testing), row.names = c("Observations", "Variables")))
```

Because there were a lot of variables for each dataset, some of incomplete varibles, which mostly contained NA, blank value or had very few unique values, were need to be removed. While, the first seven variables were only the information that might not had an major contribution to the measurement.

``` {r cleaning the data}
# Remove the first seven variables:  
training <- training[ ,-c(1:7)]
testing <- testing[,-c(1:7)] 
# Remove the incomplete variables:  
training <- training[ ,colSums(is.na(training)) == 0]
testing <- testing[ ,colSums(is.na(testing)) == 0]
# Remove the very few unique values (near zero variance):
training <- training[, -nearZeroVar(training)]
testing  <- testing[, -nearZeroVar(testing)]
# Data for processing
t(data.frame(Training = dim(training), Testing = dim(testing), row.names = c("Observations", "Variables")))
```

Outliers would caused the significant model error, therefore, it was required to verify and replace with NA value before the models were built.

``` {r removing the outliers}
strip_outliers <- function(v, upperz = 3, lowerz = -3, replace = NA) 
{
        z <- scale(v)   # gives z-scores for vector v
        v[!is.na(z) & (z>upperz | z<lowerz)] <- replace
        return(v)       # replace outliers with NA
}  
trainning <- data.frame(lapply(training[,-c(53)], strip_outliers))
```

To minimize the predictors and reduce the noise, all (52) variables of the training were analyzed and performed the correlation plot as following:  

``` {r analyze and plot the correlation}
cor.mat <- abs(cor(training[, -53]))
highlyCorrelated <- findCorrelation(cor.mat, cutoff=0.9)
names(training)[highlyCorrelated]
```

The list showed the highly correlated variables. However, as the number of correlations are quite few, PCA (Principal Components Analysis) step might not be required to perform for this assignment.  

# Model Building with cross validation  
There were 3 methods, would be applied to create the prediction models from the training dataset, while the highest model accuracy would be applied to the validation dataset, which will be also used for the quiz predictions.  
Those 3 methods were: (1) Classification Trees: CART, (2) Random Forests: RF, and (3) Generalized Boosted Regression Modeling: GBM.  
Note that a Confusion Matrix was determined at the end of each analysis to better visualize the accuracy of the models. Furthermore, the effects of overfitting could be limited, and improved the model efficiency by applying the cross-validation technique (5-folds), while the NA value was imputed by K-nearest neighbor (knn) method.

``` {r fitting the model}
trControl <- trainControl(method="cv", number=5, verboseIter = FALSE)
# Classification Trees
model_cart <- train(classe ~ ., method="rpart", preProcess="knnImpute", data=training, trControl=trControl)
# Random Forests
model_rf <- train(classe ~ ., method="rf", preProcess="knnImpute", data=training, trControl=trControl)
# Generalized Boosted Regression Modeling
model_gbm <- train(classe ~ ., method="gbm", preProcess="knnImpute", data=training, trControl=trControl)
```

# Model Testing and Out-of-sample Error 
``` {r Predicting model by using the testing data}
predCART <- predict(model_cart, newdata=testing)
cmCART <- confusionMatrix(predCART, testing$classe)
predRF <- predict(model_rf, newdata=testing)
cmRF <- confusionMatrix(predRF, testing$classe)
predGBM <- predict(model_gbm, newdata=testing)
cmGBM <- confusionMatrix(predGBM, testing$classe)
data.frame(Model = c("CART", "RF", "GBM"), Accuracy = c(cmCART$overall[1], cmRF$overall[1], cmGBM$overall[1]))
```

According to the assessment of these 3 model fits and out-of-sample error testing results, it was apparently indicated that both RF and GBM were significant better than CART model. With random forests being slightly more accurate; therefore, it would be applied to the validation data.

# Validation

``` {r Validating the selected RF model}
validMod <- predict(model_rf, newdata = validation); validMod
```

# Appendix  

*  Correlation plot
``` {r Correlation plot, fig.align = center}
diag(cor.mat) <- 0
which(cor.mat > 0.9, arr.ind = TRUE)
corrplot(cor.mat, order = "FPC", method = "color", type = "upper",
         tl.cex = 0.6, tl.col = rgb(0, 0, 0))
```

* Classification Trees Model

``` {r Classification trees chart, fig.align = center}
fancyRpartPlot(model_cart$finalModel)
```

* Accuracy of Random Forest Model

``` {r Random Forest Model Accuracy with Predictors, fig.align = center}
plot(model_rf,main="Accuracy of Random forest model by number of predictors")
```

* Accuracy of Generalized Boosted Regression Modeling
``` {r GBM Accuracy with Iterations, fig.align = center}
plot(model_gbm)
```